---
title: "Simulating the data"
output:
  html_document:
    df_print: paged
---


We are going to find the system response (acceleration) using the discrete state space model. The system response is given by:

$$
\ddot{z}(t) + 2\zeta_1 \omega_1 \dot{z}(t) + \omega_1^2 z(t) = F(t)
$$

where *z(t)* is the position of the system at time *t*; $\omega_1$ is the natural frequency of vibration; $\zeta_1$ is the damping ratio; 

```{r}
# Natural frequency of vibration (Hz)
f1 = 4
# Natural frequency of vibration (rad/s)
w1 = 2*pi*f1
# Damping ratio
z1 = 0.02
```

*F(t)* is the force. We consider *F(t)* is a Gaussian random process $F(t_k) \sim N(0,H)$:

```{r}
# Number of time steps
nt = 1000
# Sampling time step (sec)
dt = 0.05
# time vector
tt = (0:(nt-1))*dt
# Force
set.seed(123)
H = 1 # variance of Fk
Fk = rnorm(nt, mean = 0, sd = sqrt(H))
plot(tt, Fk, type = "l")
```

The dynamic equation can be written is state space form:

$$
\dot{x}(t) = A_c \cdot x(t) + B_c \cdot F(t)
$$
In discrete time, this equation is written as ($k = 1,2,\cdots,nt$):

$$
x_{k+1} = A \cdot x_k + B \cdot F_k
$$

$$
A = exp(A_c \cdot dt)
$$

```{r}
# A matrix
Ac = matrix(c(0, 1,-w1^2, -2*z1*w1), nrow = 2, byrow = T)
A1 = Matrix::expm(Ac*dt)
(A = matrix(A1@x, nrow = 2))
```

$$
B = (A - I) \cdot A_c^{-1} \cdot B_c
$$

```{r}
# B matrix
Bc = matrix(c(0, 1), nrow = 2)
B = (A - diag(2)) %*% solve(Ac) %*% Bc
```

Now we can compute the discrete states:

```{r}
xk = matrix(0.0, nrow = 2, ncol = nt)
for (r in 1:(nt-1)){
  xk[,r+1] = A %*% xk[,r] + B %*% Fk[r]
}
```

The system output, accelerations in this case, are found by mean of:

$$
y_k = \ddot{z}_k = C x_k + F_k
$$

```{r}
# C matrix
C = matrix(c(-w1^2, -2*z1*w1), nrow = 1)

# output
y = C %*% xk + Fk

# as vector
y = as.vector(y)

plot(tt, y, type = "l")
```

We add noise $e_k \sim N(0,E)$:

```{r}
E = 0.01*var(y) # the standard deviation of the noise is 10% of the standard deviation of the noise-free response
y = y + rnorm(nt, mean = 0, sd = sqrt(E))
plot(tt, y, type = "l")
```

```{r}
fs = 1/dt
ff = (0:(nt-1))*(fs/nt)
yw = fft(y)
plot(ff,abs(yw), type = "l", xlim = c(0,fs/2))
```


Finally, we save the system output:

```{r}
# write.table(y, file = "simula.txt", col.names = F, row.names = F, sep = ",")
```




